A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-02 latex/main.tex
Updates from Overleaf

\title{A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2}
\date{}

@@

\affil[*]{Correspondence: shing.zhan@bdi.ox.ac.uk}
\affil[**]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}

@@

[TODO we should mention UShER here: they state multiple times in the GitHub issues 
that they don't handle them at all, so I think we can just say this in some nice
way.]
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they are difficult to distinguish from bioinformatics errors 
and their multi-base nature presents challenges to the Markovian structure of 
the LS HMM, we masked out all gap characters as missing data during primary
inference. 
To incorporate the most important deletions into the ARG, we 
remapped data using posthoc parsimony at 75 sites involved in deletions 
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during postprocessing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known 
high frequency (Table~\ref{tab:major_dels}) and 
recurrent (Table~\ref{tab:rec_dels}) deletions.
For example, the 4 characteristic deletions associated with 
% FIXME: this is just my notation - is this something recognised more 
% widely? If not, fix to something that is?
Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
identified as having a single origin [CITATION?]
(Figure~\ref{fig:subgraph_alpha}).
The 21765+6 deletion (Spike 69-70)
which causes S-Gene Target Failure \cite{Walker2021},
emerges two more times at the origins of the 
BA.1 and BA.4 variants.
Although most highly recurrent deletions are likely errors,
some known truly recurrent deletions are present,
for example, two deletions in S1-NTD [WHICH TWO???]
associated with antibody escape\cite{McCarthy2021}.
See Document\ref{sec:supplement_dels} for further analysis.

@@

This work was funded by [FUNDER] via grant [GRANT NO.].
The authors thank all members of the lab for their support.

@@

% Data from
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_rewire-recombinants-report.ipynb
Using recombinant-rematching,
58 recombinants were identified as being more parsimoniously 
explained by matching to existing nodes in the ARG (inserted by 
subsequent parsimony improving heuristics) and 16 by long branch 
splitting. Of these 16 recombinants, two are noteworthy as 
being the origins of the Delta and BA.2 lineages in the ARG.
Both result in substantially fewer mutations (15 fewer for 
Delta, 6 for BA.2) than the previous recombinant HMM solution.
These major saltational lineages highlight the challenges 
of using the LS HMM with highly diverged samples.
We then updated the ARG to remove these 74 artefactual 
recombinants by ``rewiring'' the recombination node to the 
more parsimonious non-recombinant solution. This 
resulted in 16 additional nodes, 182 fewer mutations and reduced
the number of trees along the genome from 348 to 317. The long branches
involved in the origins of Delta and BA.2 originally joined
B.1 to B.1.167 and B.1.1 to BA.1 respectively; they can be seen
in Figure \ref{fig:subgraph_delta} and \ref{fig:subgraph_ba1_ba2}
with inserted intermediate nodes leading to the Delta and BA.2
origination nodes.

@@

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions (MNS \cite{DeMaio2020issues})
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
Fortunately, artefactual recombinants caused by this effect can be identified
because they affect multiple adjacent or near-adjacent bases simultaneously.

@@

@@

We compare the non-recombinant parts of the ARG with the UShER topology
by simplifying the phylogenies to a smaller number of
Pango representative samples.
We focus on the 1473 Pango lineages that have perfect ``origination events'' 
as described in \ref{sec:supplement_pango_lineages}, and
omit Pango designations starting with "X",
to avoid recombinants, leaving 1329 Pango lineages.
To ensure that each sample is a tip (rather than an internal node)
in the simplified topologies,
a representative sample is then chosen as the oldest sample
descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We then further remove 4 samples
which are singleton descendants of a recombination event in the ARG
(representing potentially missed Pango recombinant lineages),
as well as 43 samples that are descendants of the recombinant XBB
but which do not start with an "X".
This results in a simplified ARG of 1282 samples,
each sample having a different Pango designation.

The simplified ARG contains two single-breakpoint recombination nodes. The first
corresponds to a BA.2 recombinant whose breakpoint is on the far right (position 27382) supported by only 4 sites / 3 loci, anch which therefore does not pass our recombinant quality control measure. The second is the origin of BA.5, which (the ).
To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
and extract non-recombinant structures from the ARG
by creating a ``base tree'' of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
a Delta subtree of 160 Pangos,
a BA.2 subtree of 151 sample (excluding BA.5 descendants) and
a BA.5 subtree of 251 samples.
Figure SX shows each of these trees compared against the equivalent dated UShER topology,
with the left and right parents of Delta and BA.2 highlighted in the base tree.
The order of tips in the two trees was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
The entire pipeline is saved as a notebook at URL (?).

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed
a more comprehensive mutational landscape of SARS-CoV-2
by mapping deletions onto the ARG using post-hoc parsimony.
We identified major deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur in major VOCs 
(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
\url{https://github.com/cov-lineages/pango-designation/issues/361};
\cite{Stern2021,Akaishi2023,Li2023indels},
except the 4-base deletion and two 5-base deletions in ORF1ab,
which are likely artefacts of deletion remapping (see below).
Notably, eight of these deletions occur
in the N-terminal domain of the S1 unit in Spike,
a known hotspot for recurrent deletions \cite{McCarthy2021}
(also see Table~\ref{tab:rec_dels}).

Based on a phylogenetic analysis of 25 randomly chosen samples,
Akaishi and Fujiwara \cite{Akaishi2023} hypothesized
the emergence of commonly observed indels in SARS-CoV-2.
We found support in the ARG for the the phylogenetic placement of
these VOC-associated deletion events suggested in that study:
a 3-base deletion 21,991 above the Alpha origin node;
two 6-base deletions at 22,029 and 28,248 above the Delta origin node;
a 9-base deletion at 28,362 above the B.1.1.529 origin node;
a 3-base deletion at 6,513,
a 9-base deletion at 21,987, and
a 3-base deletion at 22,194 above the BA.1 origin node;
a 9-base deletion at 21,633 above the BA.2 origin node.
However, the ARG suggests that two major deletions
(a 6-base deletion at 21,765 and a 1-base deletion at 28,271;
Table~\ref{tab:major_dels}) are recurrent
(i.e., acquired multiple times),
but Akaishi and Fujiwara \cite{Akaishi2023} suggest that
these deletions might have been acquired
before the divergence of certain VOCs,
based on an analysis of a small set of 20 samples.
That study suggests that the 6-base deletion
occurred before the divergence of Alpha and Omicron,
but in the ARG this deletion occurs three times,
above the origin nodes of Alpha, BA.1, and BA.4.
Similarly, that study suggests that the 1-base deletion
occurred before the divergence of Alpha, Delta, and Omicron,
but in the ARG this deletion occurs
above the Alpha origin node and above node 1436808.

We identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG shown in Table~\ref{tab:rec_dels}.
These include known recurrent deletions, for example,
6-base deletion at 21,765,
3-base deletion at 21,991, and
3-base deletion 22,194 in S1-NTD
associated with antibody escape \cite{McCarthy2021,Meng2021}.

% SHZ: Talk about the agreement before getting into improvements.

There are three contiguous deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node (starting at 11,283) and
the other above the BA.2 origin node (starting at 11,288)
in the ARG (Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up during deletion remapping
as they had four overlapping sites at 11,288-11,292
(which was mapped as the 4-base deletion event).
Furthermore, Akaishi and Fujiwara \cite{Akaishi2023}
suggest that a 9-base deletion at 11,288 occurred
before the divergence of Alpha and Omicron.
In the ARG, however, this deletion occurs
above the Alpha origin node.
The out-of-frame deletions in S1-NTD
(e.g., the 2-base deletion at 21,765),
N, ORF1ab, and ORF8
are implausible and are likely artefacts caused by
sequencing errors or sequence misalignments.
Future improvements in
data pre-processing and deletion remapping
should remove such deletions.

@@

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-BA.1+BA.2.pdf}
\caption{
Subgraph of BA.1 and BA.2.
[Some notes to help the reader know this 
reflects known events]
}
\label{fig:subgraph_ba1_ba2}
\end{figure}

>>>>

\title{A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2}
\date{}

@@

\affil[*]{Correspondence: shing.zhan@ndph.ox.ac.uk}
\affil[**]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}

@@

[TODO we should mention UShER here: they state multiple times in the GitHub issues 
that they don't handle them at all, so I think we can just say this in some nice
way.]
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they are difficult to distinguish from bioinformatics errors 
and their multi-base nature presents challenges to the Markovian structure of 
the LS HMM, we masked out all gap characters as missing data during primary
inference. 
To incorporate the most important deletions into the ARG, we 
remapped data using posthoc parsimony at 75 sites involved in deletions 
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during postprocessing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known 
high frequency (Table~\ref{tab:major_dels}) and 
recurrent (Table~\ref{tab:rec_dels}) deletions.
For example, the 4 characteristic deletions associated with 
% FIXME: this is just my notation - is this something recognised more 
% widely? If not, fix to something that is?
Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
identified as having a single origin [CITATION?]
(Figure~\ref{fig:subgraph_alpha}).
The 21765+6 deletion (Spike 69-70)
which causes S-Gene Target Failure \cite{Walker2021},
emerges two more times at the origins of the 
BA.1 and BA.4 variants.
Although most highly recurrent deletions are likely errors,
some known truly recurrent deletions are present,
for example, three deletions in S1-NTD
(21,765+6, 21,991+3, and 22,194+3)
associated with antibody escape\cite{McCarthy2021}.
See Document\ref{sec:supplement_dels} for further analysis.

@@

SHZ is supported by an Intermediate Research Fellowship
from the Nuffield Department of Population Health,
University of Oxford.

@@

% Data from
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_rewire-recombinants-report.ipynb
Using recombinant-rematching,
58 recombinants were identified as being more parsimoniously 
explained by matching to existing nodes in the ARG (inserted by 
subsequent parsimony improving heuristics) and 16 by long branch 
splitting. Of these 16 recombinants, two are noteworthy as 
being the origins of the Delta (B.1.617.2) and Omicron-BA.2 lineages in the ARG.
Both result in substantially fewer mutations (15 fewer for 
Delta, 6 for BA.2) than the previous recombinant HMM solution.
These major saltational lineages highlight the challenges 
of using the LS HMM with highly diverged samples.
We then updated the ARG to remove these 74 artefactual 
recombinants by ``rewiring'' the recombination node to the 
more parsimonious non-recombinant solution. This 
resulted in 16 additional nodes, 182 fewer mutations and reduced
the number of trees along the genome from 348 to 317. The long branches
involved in the origins of Delta and BA.2 originally joined
B.1 to B.1.167 and B.1.1 to BA.1 respectively; they can be seen
in Figure \ref{fig:subgraph_delta} and \ref{fig:subgraph_omicron}
with inserted intermediate nodes leading to the Delta and BA.2
origination nodes.

@@

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions (MNS \cite{DeMaio2020issues})
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
Fortunately, artefactual recombinants caused by this effect are usually
identifiable because they affect multiple adjacent or near-adjacent
bases simultaneously.

@@

\subsubsection*{Likely spurious influential recombinants}
In the sc2ts ARG, there are two recombination events which have a large number of
descendants that are not classified as recombinants by Pango. The first of these
corresponds to recombination event within BA.2 whose breakpoint is on the far right
(position 27382) supported by only 4 sites / 3 loci, and which therefore does not
pass our recombinant quality control measure. The second of these involves a single
breakpoint at corresponds to
the origin of the BA.5 lineage

@@

We compare the non-recombinant parts of the ARG with the UShER topology
by simplifying the phylogenies to a smaller number of
Pango representative samples.
We focus on the 1473 Pango lineages that have perfect ``origination events'' 
as described in \ref{sec:supplement_pango_lineages}, and
omit Pango designations starting with "X",
to avoid recombinants, leaving 1329 Pango lineages.
To ensure that each sample is a tip (rather than an internal node)
in the simplified topologies,
a representative sample is then chosen as the oldest sample
descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We then further remove 4 lineages
which are singleton descendants of a recombination event in the ARG
(representing potentially missed Pango recombinant lineages),
as well as 43 samples that are descendants of the recombinant XBB
but which do not start with an "X".
This results in a simplified ARG of 1282 samples,
each sample having a different Pango designation.

The simplified ARG contains two single-breakpoint recombination nodes. The first
 The second is the origin of BA.5 which we also suggest may be artefactual (see Supp FIXME).
To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
and extract non-recombinant structures from the ARG
by creating a ``base tree'' of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
a Delta subtree of 160 Pangos,
a BA.2 subtree of 151 sample (excluding BA.5 descendants) and
a BA.5 subtree of 251 samples.
Figure SX shows each of these trees compared against the equivalent dated UShER topology,
with the left and right parents of Delta and BA.2 highlighted in the base tree.
The order of tips in the two trees was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
The entire pipeline is saved as a notebook at URL (?).

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed
a more comprehensive mutational landscape of SARS-CoV-2
by mapping deletions onto the ARG using post-hoc parsimony.
We identified major deletion events, for example,
the deletions inherited by at least 10,000 samples
shown in Table~\ref{tab:major_dels}.
% This just says that the deletions are found in sequences of VOCs.
All these deletions are known to occur in major VOCs 
(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
\url{https://github.com/cov-lineages/pango-designation/issues/361};
\cite{Stern2021,Tegally2022,Li2023indels}),
except the 4-base deletion and two 5-base deletions in ORF1ab,
which are likely artefacts of deletion remapping or
sequence misalignment (see below).
Notably, eight of these deletions occur
in the N-terminal domain of the S1 unit in Spike,
a known hotspot for recurrent deletions \cite{McCarthy2021}
(also see Table~\ref{tab:rec_dels}).

% Placement of deletion events linked to VOCs
In the ARG, we found that the following deletion events
are associated with the origins of major VOCs
(Table~\ref{tab:major_dels}):
a 9-base deletion at 11288,
a 6-base deletion at 21,765, and
a 3-base deletion at 21,991 above the Alpha origin node;
a 6-base deletion at 22,029 above the Delta origin node;
a 6-base deletion at 28,248 above a child of the Delta origin node;
a 9-base deletion at 28,362 above the B.1.1.529 origin node;
a 3-base deletion at 6,513,
a 3-base deletion at 22,194,
a 6-base deletion at 21,765, and
a 9-base deletion at 21,987 above the BA.1 origin node;
a 9-base deletion at 21,633 above the BA.2 origin node; and
a 6-base deletion at 21,633 above the BA.4 origin node.
All these deletion events have been proposed previously
to be linked with the emergence of these VOCs
(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
\url{https://github.com/cov-lineages/pango-designation/issues/361};
\cite{Stern2021,Tegally2022}).

% Known recurrent deletions
Using the ARG, we identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG
shown in Table~\ref{tab:rec_dels}.
These include known recurrent deletions, for example,
a 6-base deletion at 21,765,
a 3-base deletion at 21,991, and
a 3-base deletion 22,194 in S1-NTD
associated with antibody escape \cite{McCarthy2021,Meng2021},
which are also caused by major deletion events in the ARG
(Table~\ref{tab:major_dels}).

% Problems and future improvement
Detection of deletions and inference of deletion events
are difficult, especially when the dataset is enormous.
Here we used a naive procedure to map deletions onto the ARG,
treating each site involved in a deletion
as independent of the other sites.
Consequently, some deletion events might be split up and
appear as multiple smaller deletion events,
which are likely artefactual.
The three contiguous out-of-frame deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292)
are examples of such artefacts (Table~\ref{tab:major_dels}).
In the sequence alignments, there are two 9-base deletions,
one at 11,283 which commonly occur in BA.1 samples and
the other at 11,288 which commonly occur in BA.2 samples.
Assuming that the alignments are correct,
there should be two deletion events,
but they were mapped as three deletion events
because they had four overlapping sites from 11,288 to 11,292,
which were inferred by parsimony
as the 4-base deletion event at 11,288.
Furthermore, sequencing errors or sequence misalignments
can also cause artefactual deletions,
for example, the out-of-frame recurrent deletions
in N, ORF1ab, ORF8, and particularly S1-NTD
(e.g., the 2-base deletion at 21,765)
(Table~\ref{tab:rec_dels}).
Future improvements in deletion remapping and data pre-processing
should remove such artefactual deletions.

@@

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-Omicron.pdf}
\caption{
Subgraph of the BA.1 and BA.2 Omicron lineages. BA.1 samples are shown 
[Some notes to help the reader know this 
reflects known events]
}
\label{fig:subgraph_omicron}
\end{figure}
